\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-21 tex/paper.tex
save

For fixed $\theta_f$, the adversary $r$ is optimal at $$\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r),$$ in which case $p_{\hat\theta_r}(z|f(X;\theta_f)=s) =
p(z|f(X;\theta_f)=s)$ for all $z$ and all $s$, and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{s \sim f(X;\theta_f)} [ H({Z|f(X;\theta_f)=s}) ]$ of the conditional distribution of the nuisance.
This expectation is nothing else than the conditional entropy of the random variables
$Z$ and $f(X;\theta_f)$ and can be written as $H(Z|f(X;\theta_f))$.
Accordingly, the
value function $E$ can be restated as $$E'(\theta_f) = {\cal
L}_f(\theta_f) -  H({Z|f(X;\theta_f)}).$$  In particular, we have the lower
bound $$H({Y|X}) - H(Z) \leq {\cal L}_f(\theta_f) - H({Z|f(X;\theta_f)})$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $H({Y|X})$,

    \item $\smash{\hat\theta}_f$ maximizes the conditional entropy
    $H({Z|f(X;\theta_f)})$ since $H(Z|f(X;\theta)) \leq H(Z)$.
\end{itemize}
When the lower bound is active, we have $H(Z|f(X;\theta_f)) = H(Z)$
because of the second condition, which happens exactly when $Z$ and $f(X;\theta_f)$
are independent variables. In other words,  the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity.

@@

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, nodes for the standard deviations with exponential
activations to ensure positivity, while nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in the lower plots of Figure~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
% Finally, let us note that the ideal situation of a
% classifier that is both optimal and pivotal appears to be unreachable for this
% problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
% {\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H({Y|X}) -
% \lambda H(Z)]$.

>>>>

For fixed $\theta_f$, the adversary $r$ is optimal at $$\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r),$$ in which case $p_{\hat\theta_r}(z|f(X;\theta_f)=s) =
p(z|f(X;\theta_f)=s)$ for all $z$ and all $s$, and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{s \sim f(X;\theta_f)} [ H({Z|f(X;\theta_f)=s}) ]$ of the conditional distribution of the nuisance.
This expectation is nothing else than the conditional entropy of the random variables
$Z$ and $f(X;\theta_f)$ and can be written as $H(Z|f(X;\theta_f))$.
Accordingly, the
value function $E$ can be restated as a function depending on $\theta_f$ only: $$E'(\theta_f) = {\cal
L}_f(\theta_f) -  H({Z|f(X;\theta_f)}).$$  In particular, we have the lower
bound $$H({Y|X}) - H(Z) \leq {\cal L}_f(\theta_f) - H({Z|f(X;\theta_f)})$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $H({Y|X})$,

    \item $\smash{\hat\theta}_f$ maximizes the conditional entropy
    $H({Z|f(X;\theta_f)})$ since $H(Z|f(X;\theta)) \leq H(Z)$. Note that this
    latter inequality holds for both the discrete and the differential definitions of entropy.
\end{itemize}
When the lower bound is active, we have $H(Z|f(X;\theta_f)) = H(Z)$
because of the second condition, which happens exactly when $Z$ and $f(X;\theta_f)$
are independent variables. In other words,  the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity.

@@

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, nodes for the standard deviations with exponential
activations to ensure positivity, while nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in the lower plots of Figure~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $H({Y|X}) -
\lambda H(Z)$.
